Spatial considerations in the analysis of biological conservation data: space is special

### 2016-12-01 paper_v2_GEB.Rmd
Add files via upload

---
title: 'Spatial considerations in the analysis of biological conservation
data: space is special'
author: "Alexis Comber^1^, Steve Carver^1^, Carol Ximena Garzon-Lopez^2^, Paul Harris^3^, Duccio Rocchini^4^"
output: pdf_document
geometry: margin=3cm
---

@@

# Abstract
This paper demonstrates the use of spatially explicit, local statistical models to explore accommodate spatial autocorrelation or spatial non-stationarity commonly observed in statistical patterns and relationships. It applies a  geographically weighted regression (GWR) to test the model suggested by Coudun et al (2006) for the presence of *Acer campestre*. However, local collinearity amongst predictor variables may be found in local models such as GWR, even when not observed globally. The analysis therefore tests for local correlations amongst variables and applies a locally compensated GWR model where it is locally needed. In this way, this paper describes the spatial variation of coefficient estimates predicting *Acer campestre* and compensates for any local correlations amongst variables using a local ridge term. In so this paper emphasises the need for Geography Goggles in applied geographical analyses of data that increasingly have a spatial component in the form of location, for instance from GPS. It argues that space is special and encourages explicitly spatial ways of thinking. These ideas are posited in the context of nearly all data being spatial (i.e location is included), and the need for slightly more informed approaches for analysing space and location than provided by standard statistical models.

@@

This paper emphasises the importance of explicitly considering the spatial properties of data that increasingly have location attached to them, thanks to near ubiquitous GPS in measurement and observation. Space is special and location is not just another variable, precisely because of spatial autocorrelation or spatial non-stationarity, suggesting the need for  *Geography Goggles* in applied geographical analyses and spatially explicit analytical methods and models. In order to demonstrate the value of spatially explicit methods, this  paper tests the regression model suggested by Coudun et al (2006) but does this using a Geographically Weighted (GW) framework (Brunsdon et al, 1996), in this case a GW Regression (GWR). Coudun et al (2006) found the presence of *Acer campestre* in France to be significantly related to two factors rainfall and evapotranspiration. Here Coudon’s analysis is extended to account for a particular form of spatial dependence in the data which occurs when changes in properties of nearby features are found to be correlated. This contradicts the underlying assumptions of independence and stationarity made in many statistical analyses and inferences. The result is spatial autocorrelation or spatial non-stationarity when the statistical pattern or statistical relationship observed in one location differs from that in another. To explore these, this paper applies a GWR analysis to examine the spatial variation in the relationships between the predictor variables and *Acer campestre* (field maple) presence to test for the presence of local, non-stationary relationships. GWR is directly concerned with modelling spatial non-stationarity  and in doing so will often indirectly account for any spatial autocorrelation (e.g. Harris etc al. 2011)

@@

Understanding species distribution and spatial dependencies is a key concept in ecological research (Rocchini et al., 2015), in particular when dealing with biodiversity monitoring for conservation practices (Honrado et al., 2016). Several papers have dealt with the importance of explicitly taking space into account when trying to model the distribution of a certain species, and, overall, its change in space and time (see Pearman et al., 2008 and references therein). The description of the distribution of biodiversity at different spatial and temporal scales has long been the focus of ecology and biogeography. Reliable descriptions of species distributions are fundamental for conservation and research purposes (Dormann, 2007) and for conservation, where lack of accuracy hamper its potential to provide effective tools for the development of well-tailored, in space and time, biodiversity management strategies (Guisan et al., 2014, Porfirio et al., 2014). However, when modelling species spread, a number of statistical problems must be solved before spatial models can be built such as: the uncertainty related to the sampling of the modelled species (Rocchini et al., 2011), the correlation among predictors (Dormann et al., 2012), the potential variability over space of the predictors being used (Rocchini et al., 2016), the spatio-temporal dependency of the predictors (Zuur and Ieno, 2016). Among these, spatial non-stationarity and scale dependence are expected to impact most of the ecological processes and relative patterns such as the distribution of biodiversity over space (Foody, 2004).

@@

The major issue with modelling spatial data is the lack of independence of the spatial objects and the assumption of stationarity in the processes being modelled. These are particularly true with socio-economic data, but are also true of environmental and ecological data. Both concerns relate to Tobler's dictum (the first Law of Geography) that: '*everything is related to everything else, but near things are more related than distant things*' (Tobler 1970). However, in spite of numerous methodological advances in addressing such spatial modelling problems, policy related research is still routinely informed by non-spatial modelling practices, which are heroic at best, and ill-advised at worst. The lineage, from the late 1970s onwards, of such methodological advances can be loosely traced through spatial statistics texts from Journel and Huigbrechts (1978) to Cressie (1993) to Chilès and Delfiner (1999) to Cressie and Wilke (2011).

@@

Data recorded between 1980 and 2010 describing the presence of *Acer campestre* was downloaded from GBIF using the `dismo` R package, and subsetted for the UK. The data contained 22,701 records whose spatial distribution are shown in Figure 1, with a median of `r summary(colSums(res.vec))[3]`  records per year, a 1st quartile of `r summary(colSums(res.vec))[2]`  and a `r summary(colSums(res.vec))[4]` quartile of 917 records. The data for all years were summed over Ordnance Survey 10km grids because of the uneven distribution in time and space. Potential alternative approaches including generating absence points, background data (Phillips et al. 2009) to charcaterise study area environments or pseudo-absences (e.g. VanDerWal et al., 2009), indicating where absences might occur. However, pseudo absence approaches require a number of assumptions and lack statistical methods for handling the overlap between presence and background points (Ward et al. 2009; Phillips and Elith, 2011), absence data may be biased and or incomplete (Kery et al., 2010) and background data approaches generate the same measures irrespective of where the species is observed (Hijmans and Elith, 2015).

@@

The study by Coudun et al (2006) found the presence of *Acer campestre* to be significantly related to Autumn rainfall and actual Thornthwaite evapotranspiration. In this study rainfall, Potential evapotranspiration (PET) and wildness index data were used to construct a series of models for predicting the density of *Acer campestre* occurrence. Data on rainfall were downloaded from the NERC Environmental Information Data Centre (Tanguy et al, 2015) which provides monthly 1km estimated rainfall data for each year. The average Autumn (3 month) rainfall was calculated for each 1km. Mean annual PET data were downloaded from the CGIAR Consortium for Spatial Information (Trabucco and Zomer, 2009) which provides global data at 0.0083 degrees, approximately 1km. Finally, a wilderness quality dataset was included in the model. This was to explore the degree to the presence of *Acer campestre* may be related to anthropogenic disturbance – anecdotally this species is frequently used as an ornamental tree and found in field margins and hedges. The Wilderness Quality Index data were generated for the whole of Europe at 1km resolution as described in Kuiters et al (2013). WQI can be considered as measure of non-anthropogenic activity and is based on a weighted linear combination of wilderness attributes including naturalness of vegetation patterns, remoteness from settlements and human influence, and remoteness from roads. Each of these datasets were spatially aggregated over the OS 10m grid cells to generate mean values as shown in Figure 2.

@@

This study undertakes a GWR analysis, with a locally-compensated ridge term if necessary, over a 200m grid of points covering the study area, with the aim of examining the spatial distribution of coefficient estimates predicting *Acer campestre* distributions and their spatial variation. Euclidean distances were used to weight data points under the kernel. These distances better reflect the spatial processes and relationships in environmental systems than network distance (Comber et al., 2008). For the kernel, an adaptive bi-square weighting function was applied, although a number of kernel functions can be specified for GW models as discussed in Gollini et al (2015). This generates higher weights at locations very near to the kernel centre relative to those towards the edge. For each data point ($P_j$) under the kernel (with a given bandwidth), a weight $w_{i,j}$ is calculated based on its distance to the centre of the kernel ($K_{i}$) as follows:

@@

Globally, the Variance Inflation Factors (VIFs) are all less than 10, although 2 of the Variance Decomposition Proportions (VDPs) are greater than 0.5 and the Condition Number (CN) is greater than 30, using standard heuristics from Belsley et al (1980) and O’Brien (2007). Together these suggest the presence of variable collinearity when evaluated globally considering all data points together. Applying GWR collinearity diagnostics to the GWR model above generates local VIFs, local VDPs and local CNs at the same scale (i.e. using the same adaptive bandwidth of 21 data points). The results in Table 3 indicate a high degree of *local* collinearity in the GWR model. These values suggest that the application of a LCR-GWR is warranted. The local collinearity measures are mapped in Figure 5.

@@

In this paper a series of analyses were undertaken to demonstrate the application and value of explicitly spatial analyses, focusing on GWR. These develop local statistical models in order to test for spatial non-stationarity and are in contrast to standard, a-spatial, statistical approaches that assume the relationships between factors to be the same everywhere.

Additionally, the paper highlights the importance of considering and testing for local collinearity especially in spatial non-stationarity models such as GWR, even where none is found to exist globally. In this analysis very strong evidence for local collinearity was found when the data were tested using local collinearity diagnostics. In most applications of GWR this critical step is missed. Where local collinearity is found some locally-compensated models can be applied and Brunsdon et al. (2012) propose a variety of such models. A LCR_GWR only fits local ridge regressions at locations where the local CN is above a user-specified threshold. Gollini et al (2015) reviews LCR-GWR with respect to alternative approaches for handling collinearity, but whatever adapted GWR model is applied, each have the potential provide more accurate local coefficient estimates in the presence of collinearity than that found with a standard GWR model (Brunsdon et al. 2012).

The Geographically Weighted (GW) paradigm offers an attractive and coherent framework for many areas of applied geographical analysis. Geographically Weighted (GW) approaches develop local statistical models in order to test for spatial non-stationarity and are in contrast to standard, a-spatial, statistical approaches that assume the relationships between factors to be the same everywhere. They reflect what is commonly referred to as Tobler’s 1st Law of Geography (Everything is related to everything else, but near things are more related to each other) and an understanding of the world when it is viewed through ‘Geography Goggles’. These promote a vision in which the wearer is interested in how and where things vary, does not expect (statistical) relationships to be same everywhere, does not consider the world to be not normally distributed especially in space, but rather expects processes, relationships, processes, trends etc. to vary spatially and to find clusters, hotspots, coldspots, etc.

These ideas are not new: quantitative geography in 1980s identified the need to move away from the whole map statistics, particularly Stan Openshaw’s group at Newcastle and Julian Besag’s at Durham but also Luc Anselin at Arizona. Similar ideas in arose in ecology and the advent of spatial ecology. But it is important to re-state them now for a number of reasons. First, all data are spatial now (well perhaps not quite all!) but with advent of ubiquitous GPS, most records, datasets and data points have location attached to them. Second, location is not just another variable precisely because of spatial heterogeneity and spatial dependence observed in many processes, with the result that many phenomena are *not* constant or randomly distributed, as predicated by classic statistical models. Third, the need to think spatially and to and to consider the spatial dimensions in a different way is given further salience by the increased access to and use of very powerful GIS software. This is increasingly resulting in instances of poor and inappropriate use of very powerful tools, but that is another story (see Comber et al., 2015). Finally, we simply observe that geography goggles are not usually worn by researchers working in many areas of applied geography, especially conservation, environmental science and remote sensing, where the whole map statistic persists.

@@

@@

@@

@@

@@

>>>>

---
title: 'Spatial considerations in the analysis of biological conservation data: space is special'
author: "Alexis Comber^1^, Steve Carver^1^, Carol Ximena Garzon-Lopez^2^, Paul Harris^3^, Duccio Rocchini^4^"
output: 
  pdf_document: 
    keep_tex: yes
geometry: margin=3cm
---

@@

# Normal Abstract
Geographically Weighted Regression (GWR) as first proposed by Brunsdon et al (1996) is a commonly used approach in many local geographical and biogeographical analyses. One of the of criticisms of the use of GWR is that local models may be subject to local collinearity between variables, even when none is observed globally. As yet none of the published research describing applications of GWR has addressed this issue. This paper does so. It applies GWR to the model suggested by Coudun et al (2006) to predict the presence of *Acer campestre* and to account for any spatial non-stationarity, as observed in many statistical patterns and relationships. The analysis tests for presence of local collinearity amongst predictor variables and applies a locally compensated GWR model where it is needed. The results describe the spatial variation of coefficient estimates predicting *Acer campestre* and compensates for any local collinearity amongst variables using a local ridge term. The paper discusses the need for *geography goggles* in  analyses of spatial and in so doing it encourages explicitly spatial ways of thinking. Such considerations are increasingly relevant as more and more data have a spatial component in the form of location, for instance from GPS, and all data are collected some*where*. These suggest the need for more informed approaches for analysing space and location than provided by standard statistical models.

# GEB Abstract
**Aim**: The aim of this research was to test the model suggested by Coudun et al (2006) that predicts the presence of *Acer campestre*, to explore species distributions models using Geographically Weighted Regression (GWR) as first proposed by Brunsdon et al (1996), and to take account of local collinearity amongst predictor variables. Local collinearity may occur even when none is observed globally. None of the many papers describing applications of GWR have addressed this issue.\
**Location**: UK \
**Time period**: 1980 and 2010\
**Major taxa studied**: *Acer Campestre*\
**Methods**: Data on the presence of *Acer Campestre* were downloaded from GBIF, and linked to data describing  autumn rainfall, wildness quality and potential evapotranspiration (PET). These data were aggregated over OS 10km squares. The analysis compares the results of predictive models from applying a standard OLS regression, a standard GWR and a locally compensated ridge GWR. A GWR diagnostics procedures was used to test for the presence of local collinearity and where found a locally compensated GWR model was applied.\
**Results**: The results show that the presence of *Acer campestre* is significantly associated with potential evapotranspiration and Wilderness Quality Index. The GWR analyses describe the spatial variation of coefficient estimates predicting *Acer campestre* and these are contrasted with the locally compensated GWR.\
**Main conclusions**: The paper discusses the need for *geography goggles* in  analyses of spatial and in so doing it encourages explicitly spatial ways of thinking. Such considerations are increasingly relevant as more and more data have a spatial component in the form of location, for instance from GPS, and all data are collected some*where*. These suggest the need for more informed approaches for analysing space and location than provided by standard statistical models.\

\newpage

@@

Ecological data increasingly have locational attributes thanks to near ubiquitous GPS in measurement and observation devices. This paper emphasises the importance of explicitly considering the spatial properties of data and argues that location cannot be treated as just another variable, precisely because of the prevalence of spatial autocorrelation or spatial non-stationarity in data, variables and statistical relationships. Space is special and requires *geography goggles* to be worn in applied biogeographical / ecological analyses and their associated methods and models. In order to demonstrate the value of spatially explicit methods, this paper tests the regression model suggested by Coudun et al (2006) but does this using a Geographically Weighted (GW) framework (Brunsdon et al, 1996), in this case a GW Regression (GWR). Coudun et al (2006) found the presence of *Acer campestre* in France to be significantly related to two factors: rainfall and evapotranspiration. Here Coudon’s analysis is extended to account for a particular form of spatial dependence in the data which occurs when changes in properties of nearby features are found to be correlated. This contradicts the underlying assumptions of independence and stationarity made in many statistical analyses and inferences. The result is spatial autocorrelation or spatial non-stationarity when the statistical pattern or statistical relationship observed in one location differs from that in another. To explore these, this paper applies a GWR analysis to examine the spatial variation in the relationships between the predictor variables and *Acer campestre* (field maple) presence to test for the presence of local, non-stationary relationships. GWR is directly concerned with modelling spatial non-stationarity  and in doing so will often indirectly account for any spatial autocorrelation (e.g. Harris etc al. 2011)

Spatially explicit approaches such as GWR reflect Tobler's First Law of Geography (Tobler, 1970). They seek to quantify *local* patterns in relationships and processes rather than global 'whole map' statistics. Typically,  local models are constructed from subsets of the data, for example by selecting data under a moving window or kernel, and local methods such as GWR have been used in a number of biodiversity and ecological studies. Recent examples include Roll et al (2015) who examined the spatial variation tree height as a predictor of species richness for different taxa, Wang et al (2016) who used to GWR to develop local models of net primary production in Chinese forests and Keith et al (2013) who sought to predict local species richness using a GWR analysis of environmental variables. However, local methods may be subject to collinearity even when it is not observed globally (Brunsdon et al 2012) because of the particular characteristics of the local subset. Collinearity will affect model reliability, result in inferential bias through unstable parameter estimates, inflated standard errors and difficulties in separating variable effects, etc (Dormann et al 2013; Meloun et al. 2002). Collinearity is commonly tested for in analyses of a-spatial data and may even be used to predict missing data. Despite the many research papers describing applications of GWR, as yet there are no instances of research that has considered the potential effects of local collinearity. As well as extending the analysis of Coudun et al (2006) using GWR, this paper explicitly tests for and compensates for locally collinearity where found, with dramatic differences between the coefficient estimates and the mapped GWR outputs without and without local compensations. The paper proceeds with a review of approaches for handling spatial auto-correlation (Section 2), Section 3 describes data and study area along with the issues of collinearity and local collinearity, before describing the results of a GWR analysis, a GW local collinearity diagnostics analysis and a locally compensated ridge regression in Section 4. Section 5 discusses the wider implications of the research int he context of the increasing availability of spatial data and some conclusions are drawn in Section 6.

@@

Understanding species distribution and spatial dependencies is a key concept in ecological research (Rocchini et al., 2015), in particular when dealing with biodiversity monitoring for conservation practices (Honrado et al., 2016). Several papers have dealt with the importance of explicitly taking space into account when trying to model the distribution of a certain species, and, overall, its change in space and time (see Pearman et al., 2008 and references therein). The description of the distribution of biodiversity at different spatial and temporal scales has long been the focus of ecology and biogeography. Reliable descriptions of species distributions are fundamental for conservation and research purposes (Dormann, 2007) and for conservation, where a lack of accuracy hamper the potential to provide effective tools for the development of spatially and temporally informed biodiversity management strategies (Guisan et al., 2014, Porfirio et al., 2014). However, when modelling species distributions, a number of statistical problems must be solved before spatial models can be built. These include the uncertainty related to the sampling of the modelled species (Rocchini et al., 2011), correlation among predictors (Dormann et al., 2012), potential variability over space of the predictors being used (Rocchini et al., 2016) and   spatio-temporal dependencies of the predictors (Zuur and Ieno, 2016). Among these, spatial non-stationarity and scale dependence have the greatest impact on the observed ecological processes and their relative patterns, such as the distribution of biodiversity over space (Foody, 2004).

@@

The major issue with modelling spatial data is the lack of independence of the spatial objects and the assumption of stationarity in the processes being modelled. These are particularly true with socio-economic data, but are also true of bio-geographical, environmental and ecological data. Both concerns relate to Tobler's dictum (the first Law of Geography) that: '*everything is related to everything else, but near things are more related than distant things*' (Tobler 1970). However, in spite of numerous methodological advances in addressing such spatial modelling problems, policy related research is still routinely informed by non-spatial modelling practices, which are heroic at best, and ill-advised at worst. The lineage, from the late 1970s onwards, of such methodological advances can be loosely traced through spatial statistics texts from Journel and Huigbrechts (1978) to Cressie (1993), to Chilès and Delfiner (1999) and to Cressie and Wilke (2011).

@@

Data recorded between 1980 and 2010 describing the presence of *Acer campestre* was downloaded from GBIF using the `dismo` R package, and subsetted for the UK. The data contained 22,701 records whose spatial distribution are shown in Figure 1, with a median of `r summary(colSums(res.vec))[3]`  records per year, a 1st quartile of `r summary(colSums(res.vec))[2]`  and a `r summary(colSums(res.vec))[4]` quartile of 917 records. The data for all years were summed over Ordnance Survey 10km grids because of the uneven distribution in time and space. Potential alternative approaches including generating absence points, background data (Phillips et al. 2009) to charcaterise study area environments or pseudo-absences (e.g. VanDerWal et al., 2009), indicating where absences might occur. However, pseudo absence approaches require a number of assumptions and lack statistical methods for handling the overlap between presence and background points (Ward et al. 2009; Phillips and Elith, 2011), absence data may be biased and / or incomplete (Kery et al., 2010) and background data approaches generate the same measures irrespective of where the species is observed (Hijmans and Elith, 2015).

@@

The study by Coudun et al (2006) found the presence of *Acer campestre* to be significantly related to Autumn rainfall and actual Thornthwaite evapotranspiration. In this study rainfall, Potential evapotranspiration (PET) and data describing a wildness quality index were used to construct a series of models for predicting the density of *Acer campestre* occurrence. Data on rainfall were downloaded from the NERC Environmental Information Data Centre (Tanguy et al, 2015) which provides monthly 1km estimated rainfall data for each year. The average Autumn (3 month) rainfall was calculated for each 1km. Mean annual PET data were downloaded from the CGIAR Consortium for Spatial Information (Trabucco and Zomer, 2009) which provides global data at 0.0083 degrees, approximately 1km. Finally, a wilderness quality dataset was included in the model. This was to explore the degree to the presence of *Acer campestre* may be related to anthropogenic disturbance – anecdotally this species is frequently used as an ornamental tree and found in field margins and hedges. The Wilderness Quality Index (WQI) data were generated for the whole of Europe at 1km resolution as described in Kuiters et al (2013). WQI can be considered as measure of non-anthropogenic activity and is based on a weighted linear combination of attributes related to wild land, wilderness and wildness (Carver et al, 2012, Comber et al, 2010) including naturalness of vegetation patterns, remoteness from settlements and human influence, and remoteness from roads. Each of these datasets were spatially aggregated over the OS 10m grid cells to generate mean values as shown in Figure 2.

@@

This study undertook a GWR analysis, with a locally-compensated ridge term if necessary, over a 200m grid of points covering the study area, with the aim of examining the spatial distribution of coefficient estimates predicting *Acer campestre* distributions and their spatial variation. Euclidean distances were used to weight data points under the kernel. These distances better reflect the spatial processes and relationships in environmental systems than network distance (Comber et al, 2008). For the kernel, an adaptive bi-square weighting function was applied, although a number of kernel functions can be specified for GW models as discussed in Gollini et al (2015). This generates higher weights at locations very near to the kernel centre relative to those towards the edge. For each data point ($P_j$) under the kernel (with a given bandwidth), a weight $w_{i,j}$ is calculated based on its distance to the centre of the kernel ($K_{i}$) as follows:

@@

Globally, the Variance Inflation Factors (VIFs) are all less than 10, although 2 of the Variance Decomposition Proportions (VDPs) are greater than 0.5 and the Condition Number (CN) is greater than 30, using standard heuristics from Belsley et al (1980) and O’Brien (2007). Together these suggest the presence of variable collinearity when evaluated globally, considering all data points together. Applying GWR collinearity diagnostics to the GWR model above generates local VIFs, local VDPs and local CNs at the same scale (i.e. using the same adaptive bandwidth of 21 data points). The results in Table 3 indicate a high degree of *local* collinearity in the GWR model. These values suggest that the application of a LCR-GWR is warranted. The local collinearity measures are mapped in Figure 5.

@@

In this paper a series of analyses were undertaken to demonstrate the application and value of explicitly spatial analyses, focusing on GWR, and the need to reconsider common assumptions in a-spatial data analyses. Local statistical models were developed in order to test for spatial non-stationarity, in contrast to standard, a-spatial, statistical approaches that assume the relationships between factors to be the same everywhere. Additionally, the paper highlights the importance of considering and testing for local collinearity especially in spatial non-stationarity models such as GWR, even where none is found to exist globally. In this analysis very strong evidence for local collinearity was found when the data were tested using local collinearity diagnostics. In all applications of GWR published in the literature, this critical step has been missed. Where local collinearity is found, locally-compensated models such as LCR-GWR can be applied and Brunsdon et al. (2012) propose a variety of these. A LCR-GWR only fits local ridge regressions at locations where the local Condition Number is above a user-specified threshold. Gollini et al (2015) review LCR-GWR with respect to alternative approaches for handling collinearity, but whatever adapted GWR model is applied, they all have the potential provide more accurate local coefficient estimates in the presence of collinearity than that found with a standard GWR model (Brunsdon et al. 2012).

The Geographically Weighted (GW) paradigm offers an attractive and coherent framework for many areas of applied geographical analysis. Geographically Weighted approaches, testing for spatial non-stationarity, are in contrast to standard, a-spatial, statistical approaches that assume the relationships between factors to be the same everywhere. They reflect Tobler’s 1st Law of Geography and an understanding of the world when it is viewed through ‘Geography Goggles’. These promote a vision in which the wearer is interested in how and where things vary, does not expect (statistical) relationships to be same everywhere, does not consider the world to be normally distributed especially in space, but rather expects processes, relationships, processes, trends etc. to vary spatially and to find clusters, hotspots, coldspots, etc.

These ideas are not new: quantitative geography in 1980s identified the need to move away from the whole map statistics, particularly Stan Openshaw’s group at Newcastle, UK and Julian Besag’s at Durham, UK but also Luc Anselin at Arizona, USA. Similar ideas in arose in ecology and the advent of spatial ecology. But it is important to re-state them now for a number of reasons. First, all data are spatial now (well perhaps not quite all!) but with advent of ubiquitous GPS, most records, datasets and data points have location attached to them. Second, location is not just another variable precisely because of spatial heterogeneity and spatial dependence observed in many processes, with the result that many phenomena are *not* constant or randomly distributed, as predicated by classic statistical models. Third, the need to think spatially and to and to consider the spatial dimensions in a different way is given further salience by the increased access to and use of very powerful GIS software. This is increasingly resulting in instances of poor and inappropriate use of very powerful tools, but that is another story (see Comber et al., 2015). Finally, we simply observe that geography goggles are not usually worn by researchers working in many areas of applied ecology and bio-geography, especially in conservation, environmental science and remote sensing, where the whole map statistic persists.

@@

Carver S, Comber A, McMorran R and Nutter S. (2012). A GIS model for mapping spatial patterns and distribution of wild land in Scotland. *Landscape and Urban Planning*, 104 (2012) 395–409.

@@

Comber AJ, Carver S, Fritz S, McMorran R, Washtell J and Fisher P. (2010). Different methods, different wilds: evaluating alternative mappings of wildness using Fuzzy MCE and Dempster Shafer MCE. *Computers, Environment and Urban Systems*, 34: 142-152

@@

Keith SA, Kerswell AP, and Connolly SR (2014). Global diversity of marine macroalgae: environmental conditions explain less variation in the tropics. *Global ecology and Biogeography*, 23(5), 517-529.

@@

Roll U, Geffen E, and Yom-Tov Y (2015). Linking vertebrate species richness to tree canopy height on a global scale. *Global Ecology and Biogeography*, 24(7), 814-825.

@@

Wang Q, Ni J, and Tenhunen J (2005). Application of a geographically-weighted regression analysis to estimate net primary production of Chinese forest ecosystems. *Global Ecology and Biogeography*, 14(4), 379-393.
